Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors

### 2014-04-21 Direction_JINST_v12.tex
Add old versions of paper. --Lindley

\linenumbers

\abstract{Large liquid-scintillator-based detectors have proven to be
exceptionally effective for low energy neutrino measurements due to
their good energy resolution and scalability to large volumes. The
addition of directional information using Cherenkov light and fast
timing would enhance the scientific reach of these detectors,
especially for searches for neutrino-less double-beta decay. In this
paper, we develop a technique for extracting particle direction using
the difference in arrival times for Cherenkov and scintillation light,
and evaluate several detector advances in timing, photodetector
spectral response, and scintillator emission spectra that could be
used to make direction reconstruction a reality in a kiloton-scale
detector.}

@@

Scintillation light is isotropic. At these low energies, it does not contain sufficient information to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. In high pressure TPCs, reduction factors on the order of $\sim10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO,NEXTsipm}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches. 
%and merits investigation.

@@

In neutrino-electron scattering events, a single electron emerges with a distribution of energies with a maximum energy related to the incoming neutrino's energy. In comparison, $0\nu\beta\beta$ events are more complicated, with two electrons emerging with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations which depend on the underlying $0\nu\beta\beta$ decay mechanism\cite{SuperNEMO,newphysics0nuBB,bandv}. Figure 1 shows a simulated $^{116}$Cd $0\nu\beta\beta$ event in a model with light Majorana neutrino exchange, for which a probable case is the emission of two electrons with comparable energies at a large angle relative to each other.

@@

\section{Geant4 simulation}
\label{sim_section}
In order to study the effects relevant to directional reconstruction
in liquid scintillators, a Geant4\cite{geant4one,geant4two} simulation
has been constructed. The simulation uses Geant4~version 4.9.6 with the default liquid scintillator
optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. Figure \ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene and 1.52~g/l PPO. The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV),
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}.  This is a standard scintillator. The attenuation length at 400~nm, the position of the peak standard bialkali photocathode efficiency, is 25~m. The attenuation length drops precipitously between 370~nm and 360~nm from 6.5~m to 0.65~m. We use this drop to define the cutoff wavelength at 370~nm. Variations from the baseline KamLAND case are discussed below.

Re-emission of absorbed photons in the scintillator
bulk volume and optical scattering, specifically Rayleigh scattering, have not yet been included by default. A test simulation shows that the effect of optical scattering is negligible. As shown in figure~\ref{scattplot}, scattering causes a small tail at longer times. The reason is that the cutoff is very steep below 360~nm and almost no photons reach the sphere, so optical scattering makes no difference for short wavelengths. Above about 395~nm, the attenuation length is greater than 20~m so both scattering and absorption are not very likely and scattering is negligible. The intermediate region is rather small. A similar argument holds for re-emission. Scattering length measurements and discussions can be found in Ref.~\cite{Wurm:2010ad}.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs2/PEtime_Cerenkov_scattered_vs_unscattered.pdf}
        \caption[]{Cherenkov photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV). The default simulation is shown with optical scattering turned on: all photons (black solid), un-scattered photons (red dotted), and scattered (blue dashed). Scattering causes a very small tail at longer times as expected. \label{scattplot}}
        \end{center}
\end{figure}

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. As in the case of optical scattering, reflections at the sphere are a small effect that would create a small tail at longer times. Two important photodetector properties have been varied: 1)
the transit-time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}. The QE values as a function of wavelength come from the Double Chooz\cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency. We are neglecting any threshold effects in the photodetector readout electronics.

\begin{figure}
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut, 171~PEs
        from scintillation and 108~PEs from Cherenkov light are detected. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17-inch PMTs). After the 34.0~ns time cut, 349~PEs from
        scintillation and 88~PEs from Cherenkov light are detected. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut, 226~PEs from scintillation and 229~EPS
        from Cherenkov light are detected. \label{time_plots_comparison}}
        \end{center}
\end{figure}

@@

Four effects primarily contribute to the timing of the scintillator detector
system: the travel time of the particle, the time constants of the scintillation process, chromatic dispersion, and the timing of the photodetector.  First, the simulated travel time of a 5~MeV electron is 0.108$\pm$0.015~ns. This corresponds to an average path length of 3.1~cm and a final distance from the origin of 2.6~cm. The time until the electron drops below Cherenkov threshold is 0.106$\pm$0.015~ns. We note that due to scattering the final direction of the electron before it stops does not correspond to the initial direction; however the scattering angle is small while the majority of Cherenkov light is produced. The Cherenkov light thus encodes the direction of the primary electron. The scattering physics is handled by Geant4's ``Multiple Scattering" process which is valid down to 1~keV, where atomic shell structure becomes important\cite{geant4scatt}. In the energy range important for $0\nu\beta\beta$, a 1.4~MeV electron travels a total path length of 0.8~cm, has a distance from the origin of 0.6~cm in 0.030$\pm$0.004~ns  and takes 0.028$\pm$0.004~ns to drop below Cherenkov threshold. The scattering follows the same pattern.

@@

In sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations at 5~MeV. We focus
on the idea of increasing the discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the Geant4~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction. With the successful reconstruction at 5~MeV, we then lower the energy of the simulated electrons and show that it is possible to reconstruct electrons in the range interesting for $0\nu\beta\beta$.

\section{Detector timing}
\label{detector_timing_sec}

@@

In order to compare simulations with different parameters, a fixed
time cut of $t\leq$ 34.0~ns is applied using the truth information to
isolate the Cherenkov light in this early time window, as shown in figure \ref{time_plots_comparison} (a). For the default
simulation case, the average number of PEs per event coming from
Cherenkov light in the early time window (108) is 98\% of the total
average number of PEs from Cherenkov light (110). For scintillation
light, the average number of PEs (171) is only 3.1\% of the average total
scintillation-induced PEs (5445). This demonstrates the effectiveness of
a time cut to separate Cherenkov light from scintillation light.

@@

Figure \ref{Cherenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits in the early time window. The Cherenkov ring structure
can be clearly seen in the peak near 46\textdegree, demonstrating
that the directional signal conveyed by the Cherenkov photons is not
erased by scattering of the initial electrons even at 1.4~MeV.

@@

\section{Detector wavelength response}
\label{detector_wavelength_response_sec} 
Since Cherenkov photons that pass through meters of
scintillator have on average longer wavelengths than scintillation
photons, a photodetector that is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs. We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63)\cite{Hamamatsu_R3899U}, but with the default TTS of 0.1~ns.
Figure \ref{time_plots_comparison}(c) shows the results for the
modified simulation with high QE in the red spectral region. The
higher absolute number of photoelectrons coming from Cherenkov light
(factor of $\approx$ 2) and the increased Cherenkov/scintillation
ratio ($R_{C/S}=1.01$) in the early time window would significantly improve the
directionality reconstruction.

\section{Scintillator emission spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments\cite{qdot,qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectrum by tuning the size or composition of the quantum
dots.

@@

\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq2_5p0.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_5p0.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq2_2p1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_2p1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq2_1p4.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_1p4.pdf}}
        \caption[]{\label{fig:reco} (Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons. In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        at the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation. From Top to Bottom, 5~MeV, 2.1~MeV and 1.4 MeV are shown.}
\end{center}
\end{figure}

\section{Reconstruction}
\label{reconstruction_sec}

The timing studies show that in the early time window, $t\leq$
34.0~ns, the ratio $R_{C/S}$ is high, improving the photoelectron hit selection. In this section, we apply
reconstruction tools for a water Cherenkov detector, WCSimAnalysis,
to the problem of reconstructing the position and direction of 5~MeV
electrons from this early light. WCSimAnalysis is a water Cherenkov
reconstruction package developed by the Long Baseline Neutrino
Experiment (LBNE) Collaboration\cite{Blake}. It provides a framework
for generic event cleaning, track reconstruction, and particle
identification, and comes equipped with variety of pre-built
algorithms. It is continuing to be expanded using new track-fitting
techniques for water Cherenkov detectors\cite{Sanchez2012525} based on
advanced photosensors with sub-cm imaging capabilities and timing
resolutions below 100 picoseconds.

@@

For the purpose of testing the reconstruction algorithm we use 1000
simulated electrons with an energy of 5~MeV; lower energies are studied in the next section.
The electrons are simulated at the center of the detector, $\vec{r}$ = (0,0,0), along
the x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0). Figure \ref{fig:reco} (Top)
shows the vertex reconstruction. The vertex is reasonably well-reconstructed around the center of the detector, $\vec{r}$ = (0,0,0),
except along the x-axis. The RMS values of the distributions for all three
reconstructed coordinates are smaller than 3.5~cm. The shift along
the x-axis is due to two effects for which the reconstruction has to
use average values rather than the unknown true value for each
hit: the wavelength and hence the speed of the light in the medium,
and the point of emission for each of the photons, reconstructed
as coming from a common point. The reconstruction of the direction
also is shown in figure \ref{fig:reco} (Top). It shows that for the majority
of the events the initial electron direction is reconstructed well.
This is a promising result given the simplicity of the algorithms.

\section{Energy dependence}
\label{edep_section}
In the previous sections we presented results on single 5~MeV electrons such as might be observed in neutrino-electron scattering.
In this section we study two lower energies, 1.4~MeV and 2.1~MeV, appropriate for searches for $0\nu\beta\beta$. These
energies correspond to Q/2 for the double-beta decay of
$^{116}$Cd and $^{48}$Ca, respectively\cite{cd1,biller}. The isotope
$^{116}$Cd was chosen because of its potential use in quantum-dot-doped
scintillators\cite{qdot,qdot2} and $^{48}$Ca was chosen
to cover the Q-value range of $0\nu\beta\beta$ candidate isotopes.

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs/hCos_vs_E_final.pdf}
        \caption[]{Mean cosine of the angle between the true initial electron direction and the reconstructed direction, as a function of the electron energy. For each energy 1000 events have been simulated. Statistical errors are
shown. \label{Edep_angle}}
        \end{center}
\end{figure}

The reconstruction algorithms outlined in section \ref{reconstruction_sec} have also been applied
to the simulations at lower energies. Figure \ref{fig:reco} (Middle) shows the results for 2.1~MeV and figure \ref{fig:reco} (Bottom) shows the results for 1.4~MeV. Most events are still reconstructed well, despite the lower
number of PEs and the decreased $R_{C/S}$. For 1.4~MeV electrons, the RMS values of the distributions for all three
reconstructed coordinates are smaller than 4.5~cm. The mean cosine of the angle between the true direction
and the reconstructed direction for different energies of the initial electron is shown in figure
\ref{Edep_angle}. The direction reconstruction performance is still promising for energies as low as 1.4~MeV.

>>>>

%\linenumbers

\abstract{Large liquid-scintillator-based detectors have proven to be
exceptionally effective for low energy neutrino measurements due to
their good energy resolution and scalability to large volumes. The
addition of directional information using Cherenkov light and fast
timing would enhance the scientific reach of these detectors,
especially for searches for neutrino-less double-beta decay. In this
paper, we propose a technique for extracting particle direction using
the difference in arrival times for Cherenkov and scintillation light,
and evaluate several detector advances in timing, photodetector
spectral response, and scintillator emission spectra that could be
used to make direction reconstruction a reality in a kiloton-scale
detector.}

@@

Scintillation light is isotropic. At these low energies, it does not contain sufficient information to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. High pressure TPCs are another technology that can record the direction of particles. In this type of experiment reduction factors on the order of $\sim10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO,NEXTsipm}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches. 
%and merits investigation.

@@

In neutrino-electron scattering events, a single electron emerges with a distribution of energies with a maximum energy related to the incoming neutrino's energy. In comparison, $0\nu\beta\beta$ events are more complicated, with two electrons emerging with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations which depend on the underlying $0\nu\beta\beta$ decay mechanism\cite{SuperNEMO,newphysics0nuBB,bandv}. Figure 1 shows a simulated $^{116}$Cd $0\nu\beta\beta$ event in a model with light Majorana neutrino exchange, for which a probable case is the emission of two electrons with comparable energies at a large angle relative to each other. In the $0\nu\beta\beta$ decay of $^{116}$Cd, 52.1\% of the electrons have energies above 1.4~MeV and 97.4\% have energies above the Cherenkov threshold of 0.2~MeV\cite{phasespace}.

@@

\section{Geant4 simulation}
\label{sim_section}
In order to study the effects relevant to directional reconstruction
in liquid scintillators, a Geant4\cite{geant4one,geant4two} Monte Carlo (MC)  simulation
has been constructed. The simulation uses Geant4~version 4.9.6 with the default liquid scintillator
optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. Figure \ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene and 1.52~g/l PPO. The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV),
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}.  This is a standard scintillator. The attenuation length at 400~nm, the position of the peak standard bialkali photocathode efficiency, is 25~m. The attenuation length drops precipitously between 370~nm and 360~nm from 6.5~m to 0.65~m. We use this drop to define the cutoff wavelength at 370~nm. The baseline simulation is summarized in table~\ref{summary_table}. Variations from the baseline KamLAND case are discussed below.

\begin{table}
\caption{The basic parameters of the default simulation. The photodetector transit time spread (TTS), photodetector peak sensitivity, and scintillator emission spectrum are varied as part of the study. }
  \begin{center}
    \begin{tabular}{ll}
      Parameter & Value\\
      \hline\hline\\[-5px]
      Radius & 6.5~m \\
      Scintillator density & 0.78 g/ml \\
      Scintillator rise time & 1.0~ns\\
      Scintillator decay time & 6.9~ns and 8.8~ns \\
      Scintillator light yield & 9030 photons/MeV \\
      Scintillator peak emission & 400~nm \\
      Photodetector TTS & 0.1~ns \\
      Photodetector peak sensitivity & 400~nm \\
      \hline
    \label{summary_table}
    \end{tabular}
  \end{center}
\end{table}

Re-emission of absorbed photons in the scintillator
bulk volume and optical scattering, specifically Rayleigh scattering, have not yet been included by default. A test simulation shows that the effect of optical scattering is negligible. As shown in figure~\ref{scattplot}, scattering causes a small tail at longer times. The reason is that the cutoff is very steep below 360~nm and almost no photons reach the sphere, so optical scattering makes no difference for short wavelengths. Above about 395~nm, the attenuation length is greater than 20~m so both scattering and absorption are not very likely and scattering is negligible. The intermediate region is rather small. A similar argument holds for re-emission. These are second order effects that will be studied in future work on uniformly distributed and lower energy events. Scattering length measurements and discussions can be found in Ref.~\cite{Wurm:2010ad}.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs2/PEtime_Cerenkov_scattered_vs_unscattered.pdf}
        \caption[]{Cherenkov photoelectron (PE) arrival times after application
        of the photodetector transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV). The default simulation is shown with optical scattering turned on: all photons (black solid), un-scattered photons (red dotted), and scattered (blue dashed). Scattering causes a very small tail at longer times as expected. \label{scattplot}}
        \end{center}
\end{figure}

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. A uniform reduction in photodetector coverage would lower the number of measured photons. As in the case of optical scattering, reflections at the sphere are a small effect that would create a small tail at longer times. Two important photodetector properties have been varied: 1)
the transit-time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}. The QE values as a function of wavelength come from the Double Chooz\cite{dctwo}
MC simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency. We are neglecting any threshold effects in the photodetector readout electronics.

@@

Four effects primarily contribute to the timing of the scintillator detector
system: the travel time of the particle, the time constants of the scintillation process, chromatic dispersion, and the timing of the photodetector.  First, the simulated travel time of a 5~MeV electron is 0.108$\pm$0.015~ns. This corresponds to an average path length of 3.1~cm and a final distance from the origin of 2.6~cm. The time until the electron drops below Cherenkov threshold is 0.106$\pm$0.015~ns. We note that due to Coulomb scattering the final direction of the electron before it stops does not correspond to the initial direction; however the scattering angle is small while the majority of Cherenkov light is produced. The Cherenkov light thus encodes the direction of the primary electron. The scattering physics is handled by Geant4's ``Multiple Scattering" process which is valid down to 1~keV, where atomic shell structure becomes important\cite{geant4scatt}. In the energy range important for $0\nu\beta\beta$, a 1.4~MeV electron travels a total path length of 0.8~cm, travels a distance from the origin of 0.6~cm in 0.030$\pm$0.004~ns  and takes 0.028$\pm$0.004~ns to drop below Cherenkov threshold. The scattering follows the same pattern.

@@

In sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations at 5~MeV. We focus
on the idea of increasing the discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the Geant4~simulation are the photoelectron hit
positions and the detection times. The hit times are then smeared according to the TTS resolution. In section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction. With the successful reconstruction at 5~MeV, we then lower the energy of the simulated electrons and show that it is possible to reconstruct electrons in the range interesting for $0\nu\beta\beta$.

\section{Instrumentation Requirements}
\subsection{Photodetector timing}
\label{detector_timing_sec}

@@

\begin{figure}
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the photodetector transit-time spread (TTS) for the simulation
        of 1000 single electrons (5~MeV) moving along the x-axis with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut, 171~PEs
        from scintillation and 108~PEs from Cherenkov light are detected. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17-inch PMTs). After the 34.0~ns time cut, 349~PEs from
        scintillation and 88~PEs from Cherenkov light are detected. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut, 226~PEs from scintillation and 229~PEs
        from Cherenkov light are detected. \label{time_plots_comparison}}
        \end{center}
\end{figure}

In order to compare simulations with different parameters, a fixed
time cut of $t\leq$ 34.0~ns is applied using the MC truth information to
isolate the Cherenkov light in this early time window, as shown in figure \ref{time_plots_comparison} (a). For the default
simulation case, the average number of PEs per event coming from
Cherenkov light in the early time window (108) is 98\% of the total
average number of PEs from Cherenkov light (110). For scintillation
light, the average number of PEs (171) is only 3.1\% of the average total
scintillation-induced PEs (5445). This demonstrates the effectiveness of
a time cut to separate Cherenkov light from scintillation light.

@@

Figure \ref{Cherenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits in the early time window. The Cherenkov ring structure
can be clearly seen in the peak near 46\textdegree, demonstrating
that the directional signal conveyed by the Cherenkov photons is not
erased by scattering of the initial electrons even at 1.4~MeV. At this lower energy, photon statistics are reduced, making reconstruction more difficult as we move to lower energies that will occur in $0\nu\beta\beta$ and neutrino scattering events.

@@

\subsection{Photodetector wavelength response}
\label{detector_wavelength_response_sec} 
Since Cherenkov photons that pass through meters of
scintillator have on average longer wavelengths than scintillation
photons, a photodetector that is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs. We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63)\cite{Hamamatsu_R3899U}, but with the default TTS of 0.1~ns.
Figure \ref{time_plots_comparison}(c) shows the results for the
modified simulation with high QE in the red spectral region. The
higher absolute number of photoelectrons coming from Cherenkov light
(factor of $\approx$ 2) and the increased Cherenkov/scintillation
ratio ($R_{C/S}=1.01$) in the early time window would significantly improve the
directionality reconstruction.

\subsection{Scintillator emission spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments\cite{qdot,qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectrum by tuning the size or composition of the quantum
dots.

@@

\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq3_5p0.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_5p0.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq3_2p1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_2p1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq3_1p4.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_1p4.pdf}}
        \caption[]{\label{fig:reco} (Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons. In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        at the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation. From Top to Bottom, 5~MeV, 2.1~MeV and 1.4 MeV are shown.}
\end{center}
\end{figure}

\section{Event Reconstruction}
\label{reconstruction_sec}

\subsection{Description of algorithm}
The timing studies show that in the early time window, $t\leq$
34.0~ns, the ratio $R_{C/S}$ is high, improving the photoelectron hit selection. In this section, we apply
reconstruction tools for a water Cherenkov detector, WCSimAnalysis,
to the problem of reconstructing the position and direction of 5~MeV
electrons from this early light. WCSimAnalysis is a water Cherenkov
reconstruction package developed by the Long Baseline Neutrino
Experiment (LBNE) Collaboration\cite{Blake}. It provides a framework
for generic event cleaning, track reconstruction, and particle
identification, and comes equipped with variety of pre-built
algorithms. It is continuing to be expanded using new track-fitting
techniques for water Cherenkov detectors\cite{Sanchez2012525} based on
advanced photosensors with sub-cm imaging capabilities and timing
resolutions below 100 picoseconds.

@@

\subsection{Reconstruction Results}

For the purpose of testing the reconstruction algorithm we use 1000
simulated electrons with an energy of 5~MeV; lower energies are studied in the next section.
The electrons are simulated at the center of the detector, $\vec{r}$ = (0,0,0), along
the x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0). Figure \ref{fig:reco} (Left)
shows the vertex reconstruction. The vertex is reasonably well-reconstructed around the center of the detector, $\vec{r}$ = (0,0,0),
except along the x-axis. The RMS values of the distributions for all three
reconstructed coordinates are smaller than 3.5~cm. The shift along
the x-axis is due to two effects for which the reconstruction has to
use average values rather than the unknown true value for each
hit: the wavelength and hence the speed of the light in the medium,
and the point of emission for each of the photons, reconstructed
as coming from a common point. The reconstruction of the direction
also is shown in figure \ref{fig:reco} (Right). It shows that for the majority
of the events the initial electron direction is reconstructed well.
This is a promising result given the simplicity of the algorithms.

\subsection{Energy dependence}
\label{edep_section}
In the previous sections we presented results on single 5~MeV electrons such as might be observed in neutrino-electron scattering.
In this section we study two lower energies, 1.4~MeV and 2.1~MeV, appropriate for searches for $0\nu\beta\beta$. These
energies correspond to Q/2 for the double-beta decay of
$^{116}$Cd and $^{48}$Ca, respectively\cite{cd1,biller}. The isotope
$^{116}$Cd was chosen because of its potential use in quantum-dot-doped
scintillators\cite{qdot,qdot2} and $^{48}$Ca was chosen
to cover the Q-value range of $0\nu\beta\beta$ candidate isotopes.

@@

%\begin{figure}
%        \begin{center}
%        \includegraphics[scale=0.4]{graphs/hCos_vs_E_final.pdf}
%        \caption[]{Mean cosine of the angle between the true initial electron direction and the reconstructed direction, as a function of the electron energy. For each energy 1000 events have been simulated. Statistical errors are
%shown. \label{Edep_angle}}
%        \end{center}
%\end{figure}

The reconstruction algorithms outlined in section \ref{reconstruction_sec} have also been applied
to the simulations at lower energies. Figure \ref{fig:reco} (Middle) shows the results for 2.1~MeV and figure \ref{fig:reco} (Bottom) shows the results for 1.4~MeV. Most events are still reconstructed well, despite the lower
number of PEs and the decreased $R_{C/S}$. For 1.4~MeV electrons, the RMS values of the distributions for all three
reconstructed coordinates are smaller than 4.5~cm. The direction reconstruction performance is still promising for energies as low as 1.4~MeV.
